---
title: "MA678 Final Project"
author: "Yin Xu"
date: "2022-11-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, warning=FALSE, message=FALSE, include=FALSE}
library(ggplot2)
library(ggthemes)
library(ggtext)
library(cowplot)
library(Matrix)
library(lme4)
library(pbkrtest)
library(tidyr)
library(dplyr)
library(corrplot)
library(PerformanceAnalytics)
library(lmerTest)
library(lattice)
```


## Abstract

In most universities, AI, Statistics and Data Science are some of the most popular majors, plenty of students choose these majors since if they have the Data Science background, they may find jobs easily after university. According to this issue, there is a dataset from data science jobs shows that the titles, locations and other variables which effect the salaries. To discuss this topic, I build a multilevel model with group `title` and `race` The report includes 4 main parts: introduction, method, result and discussion.

## Introduction

As we always know, the job titles, locations of companies, experience level and other factors usually influence the amount of a person's salary. However, after COVID-19 came into everyone's life, some of jobs became remote to avoid direct contact, therefore, remote jobs may become one factor which influence the salaries. Remote work may decrease the salaries, because the efficiency will be reduced; sometimes, because of the extension of the work hour, the salaries will be increased. Besides, some small companies prefer to attract talents with high wages; at the same time, large companies have complete talents and do not care about the whereabouts of employees with low salaries.

Therefore, multilevel model is needed to discover the effect of the factors. I divide the factors to two group: fixed effect factors(e.g. education, bonus, location) and random effect factors(title, race).

\newpage

## Method

### Data Cleaning

The data is from kaggle open dataset(https://www.kaggle.com/datasets/jackogozaly/data-science-and-stem-salaries).

This data includes the variables of salaries from 2020 to 2021 around the world. 

First of all, there are "NA"s in the data, cleaning data is essential. Then I want to focus on the jobs in the United States, so the observations outside of USA should be dropped. Besides, I dropped several observations with super high wages (`total yearly compensation`) which is not fit for the plots. 

Secondly, there are 15 unique titles in `title`, but I only want to keep the position of the titles (e.g. manager, engineer, sales and so on). Therefore, I extracted the last word from `title` and added a new column named `title_1`, this column will be used as title for further analysis.

Additionally, in this data, `gender` and `Education` are two variables in characters, to make my exploratory data analysis more reasonable and easy to read, I changed `gender` to numeric values 0, 1, 2 which means other, female and male; for `Education`, 0 means "High School", 1 means "Some College", 2 means "Bachelor's Degree", 3 means "Master's Degree", and 4 means "PhD".

```{r include=FALSE}
salaries <- read.csv("Levels_Fyi_Salary_Data.csv")
salary <- salaries %>% drop_na()
```

```{r, warning=FALSE, include=FALSE}
#unique(salary$title)
salary$title_1 <- gsub(".+\\s(.+)$", "\\1", salary$title)
salary$state <- gsub(".+\\s(.+)$", "\\1", salary$location)
#drop

# only USA & <= 1,000,000
salary <- salary %>% 
  filter(!grepl("[a-z]", state)) %>%
  filter(totalyearlycompensation <= 1000000) %>%
  separate(timestamp, c("timestamp", "time"), sep = " (?=[^ ]+$)")

salary$timestamp <- as.Date(salary$timestamp, "%m/%d/%Y")

salary <- salary[-c(4, 10, 15, 17:28)]

salary$Education[salary$Education == "Highschool"] <- 0
salary$Education[salary$Education == "Some College"] <- 1
salary$Education[salary$Education == "Bachelor's Degree"] <- 2
salary$Education[salary$Education == "Master's Degree"] <- 3
salary$Education[salary$Education == "PhD"] <- 4
salary$Education <- as.numeric(salary$Education)

salary$gender[salary$gender == "Other"] <- 0
salary$gender[salary$gender == "Female"] <- 1
salary$gender[salary$gender == "Male"] <- 2
salary$gender <- as.numeric(salary$gender)
```

After data cleaning, the following is the chart of variables which are used in this report:

| Variables Names           | Explanation                                                                             |
| :--:                      | :-----                                                                                  |
| `timestamp`               | When the data was recorded.                                                             |
| `company`                 | Company names.                                                                          |
| `title`                   | Job title.                                                                              |
| `location`                | Job location.                                                                           |
| `cityid`                  | City ID of the location.                                                                |
| `totalyearlycompensation` | Total yearly compensation which is added by `basesalary`, `stockgrantvalue`and `bonus`. |
| `basesalary`              | Base salary of the year.                                                                |
| `stockgrantvalue`         | Stock grant value.                                                                      |
| `bonus`                   | Bonus of the year.                                                                      |
| `yearsofexperience`       | Year of experience of Data Science jobs.                                                |
| `yearsatcompany`          | Year of experience at said company.                                                     |
| `gender`                  | Gender of the observations.                                                             |
| `race`                    | Race of the observations.                                                               |
| `Education`               | Education background.                                                                   |

\newpage

### Exploratory Data Analysis

After data cleaning, I got 16937 observations and 14 variables, and I set `total yearly compensation` as the output of the analysis; the other 13 variables will be discussed in following analysis to know if they influence the salaries of these Data Science jobs. To figure out the relationships between the variables and build a model, I select `title` and `race` as two groups.

```{r, echo=FALSE, fig.height=3, fig.width=10, fig.cap="Salary vs Years at Company"}
# Years at the Company
p1 <- ggplot(data = salary) + 
  aes(x = yearsatcompany, y = totalyearlycompensation) + 
  geom_jitter(aes(color = title_1), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = title_1), method = "lm", se = FALSE, formula = 'y ~ x') + 
  ggtitle("<span style='font-size: 10pt;'>(a) Salary vs Year at the Company</font>") +
  theme(plot.title = element_markdown(), 
        text = element_text(size = 8)) +
  labs(x = "Years at the Company", y = "Total Yearly Compensation") 

p2 <- ggplot(data = salary) + 
  aes(x = yearsatcompany, y = totalyearlycompensation) + 
  geom_jitter(aes(color = Race), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = Race), method = "lm", se = FALSE, formula = 'y ~ x') + 
  ggtitle("<span style='font-size: 10pt;'>(b) Salary vs Year at the Company</font>") +
  theme(plot.title = element_markdown(), 
        text = element_text(size = 8)) + 
  labs(x = "Years at the Company", y = "Total Yearly Compensation")

plot_grid(p1, p2, labels = NULL)
```

These two graphs show the relationship between `total yearly compensation` and `years at company` which is varied by `title` and `race`. Both of the graphs show that as the years at the company increasing, the salaries increase, and the slopes are lightly different. By the first graph, the interprets of the different titles are different, which means that the very first salary after getting the Data Science jobs depends on the title, but not race(b). Besides, when I use `year of experience`, `location`, `timestamp`, `stock grant value` and `bonus` to compare the relationships with `total yearly compensation`, they showed the similar trend.

```{r, echo=FALSE, fig.height=3, fig.width=8, fig.cap="Salary vs Years at Company"}
# Education
ggplot(data = salary) + 
  aes(x = Education, y = totalyearlycompensation) + 
  geom_jitter(aes(color = title_1), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = title_1), method = "lm", se = FALSE, formula = 'y ~ x') + 
  ggtitle("<span style='font-size: 10pt;'>(a) Salary vs Education</font>") +
  theme(plot.title = element_markdown(), 
        text = element_text(size = 8)) +
  labs(x = "Education", y = "Total Yearly Compensation")

ggplot(data = salary) + 
  aes(x = Education, y = totalyearlycompensation) + 
  geom_jitter(aes(color = Race), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = Race), method = "lm", se = FALSE, formula = 'y ~ x') + 
  ggtitle("<span style='font-size: 10pt;'>(b) Salary vs Education</font>") +
  theme(plot.title = element_markdown(), 
        text = element_text(size = 8)) +
  labs(x = "Education", y = "Total Yearly Compensation")
```

According to the salary and education graphs, as the education background increasing, the slopes of most titles are increasing, there only "Recruiter" line shows the different trend. Additionally, "Analyst" and "Consultant" have no High School background, and most engineers have bachelor or master background. Overall, `title` influences `Education`, however, there is no difference of education background among diverse races.

```{r, echo=FALSE, fig.height=3, fig.width=8, fig.cap="Salary vs Gender"} 
# Gender
ggplot(data = salary) + 
  aes(x = gender, y = totalyearlycompensation) + 
  geom_jitter(aes(color = title_1), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = title_1), method = "lm", se = FALSE, formula = 'y ~ x') +
  ggtitle("<span style='font-size: 10pt;'>(a) Salary vs Gender</font>") +
  theme(plot.title = element_markdown(), 
        text = element_text(size = 8)) +
  labs(x = "Gender", y = "Total Yearly Compensation")

ggplot(data = salary) + 
  aes(x = gender, y = totalyearlycompensation) + 
  geom_jitter(aes(color = Race), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = Race), method = "lm", se = FALSE, formula = 'y ~ x') + 
  ggtitle("<span style='font-size: 10pt;'>(b) Salary vs Gender</font>") +
  theme(plot.title = element_markdown(), 
        text = element_text(size = 8)) +
  labs(x = "Gender", y = "Total Yearly Compensation")
```




## Correlation

```{r, echo=FALSE, include=FALSE}
cor <- salary[, c(7:11, 13)]
cor$yearsofexperience <- log(salary$yearsofexperience+1)
cor$yearsatcompany <- log(salary$yearsatcompany+1)
cor$basesalary <- log(salary$basesalary+1)
cor$stockgrantvalue <- log(salary$stockgrantvalue+1)
cor$bonus <- log(salary$bonus+1)
cor$gender <- log(salary$gender + 1)
cor$cityid <- log(salary$cityid+1)
cor$education <- salary$Education
cor$title <- salary$title_1
cor$race <- salary$Race
cor$salary <- log(salary$totalyearlycompensation + 1)
```

```{r, echo=FALSE, warning=FALSE, fig.cap="Correlation Matrix"}
chart.Correlation(cor[,1:7], histogram=TRUE, pch="+")
```


## Multilevel Model

```{r, warning=FALSE}
model_1 <- lmer(salary ~ yearsofexperience + yearsatcompany + cityid + education + gender
              + (1 + yearsofexperience + yearsatcompany + cityid + education | title) 
              + (1 | race), 
              data = cor, REML = F)
```

```{r include=FALSE}
summary(model_1)
anova(model_1)
```

## Model checking
```{r include=FALSE}
qqmath(model_1)
ranef(model_1)
```

\newpage


## Result



\newpage

## Appendix

```{r, echo=FALSE, fig.height=3, fig.width=8, fig.cap="Salary vs Experience Years"}
# Year of experience
ggplot(data = salary) + 
  aes(x = yearsofexperience, y = totalyearlycompensation) + 
  geom_jitter(aes(color = title_1), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = title_1), method = "lm", se = FALSE, formula = 'y ~ x') + 
  ggtitle("<span style='font-size: 10pt;'>Salary vs Year of Experience</font>") +
  theme(plot.title = element_markdown(), 
        text = element_text(size = 8)) +
  labs(x = "Experience Years", y = "Total Yearly Compensation") 

ggplot(data = salary) + 
  aes(x = yearsofexperience, y = totalyearlycompensation) + 
  geom_jitter(aes(color = Race), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = Race), method = "lm", se = FALSE, formula = 'y ~ x') + 
  ggtitle("<span style='font-size: 10pt;'>Salary vs Year of Experience</font>") +
  theme(plot.title = element_markdown(), 
        text = element_text(size = 8)) +
  labs(x = "Experience Years", y = "Total Yearly Compensation")
```

```{r echo=FALSE, fig.height=3, fig.width=8, fig.cap="Salary vs Location"}
# Location
ggplot(data = salary) + 
  aes(x = cityid, y = totalyearlycompensation) + 
  geom_jitter(aes(color = title_1), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = title_1), method = "lm", se = FALSE, formula = 'y ~ x') + 
  ggtitle("<span style='font-size: 10pt;'>Salary vs Location</font>") +
  theme(plot.title = element_markdown(), 
        text = element_text(size = 8)) +
  labs(x = "City in USA", y = "Total Yearly Compensation")

ggplot(data = salary) + 
  aes(x = cityid, y = totalyearlycompensation) + 
  geom_jitter(aes(color = Race), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = Race), method = "lm", se = FALSE, formula = 'y ~ x') + 
  ggtitle("<span style='font-size: 10pt;'>Salary vs Location</font>") +
  theme(plot.title = element_markdown(), 
        text = element_text(size = 8)) +
  labs(x = "City in USA", y = "Total Yearly Compensation")
```

```{r echo=FALSE, fig.height=3, fig.width=8, fig.cap="Salary vs Time"}
# Time stamp
ggplot(data = salary) + 
  aes(x = timestamp, y = totalyearlycompensation) + 
  geom_point(aes(color = title_1), size = .6) + 
  geom_smooth(aes(color = title_1), method = "lm", se = FALSE, formula = 'y ~ x') +
  ggtitle("<span style='font-size: 10pt;'>Salary vs Time</font>") +
  theme(plot.title = element_markdown(), 
        text = element_text(size = 8)) +
  labs(x = "Time stamp", y = "Total Yearly Compensation")

ggplot(data = salary) + 
  aes(x = timestamp, y = totalyearlycompensation) + 
  geom_point(aes(color = Race), size = .6) + 
  geom_smooth(aes(color = Race), method = "lm", se = FALSE, formula = 'y ~ x') + 
  ggtitle("<span style='font-size: 10pt;'>Salary vs Time</font>") +
  theme(plot.title = element_markdown(), 
        text = element_text(size = 8)) +
  labs(x = "Time stamp", y = "Total Yearly Compensation")
```

```{r echo=FALSE, fig.height=3, fig.width=8, fig.cap="Salary vs Stock"}
# Stock Grant Value
ggplot(data = salary) + 
  aes(x = stockgrantvalue, y = totalyearlycompensation) + 
  geom_point(aes(color = title_1), size = .6) + 
  geom_smooth(aes(color = title_1), method = "lm", se = FALSE, formula = 'y ~ x') + 
  ggtitle("<span style='font-size: 10pt;'>Salary vs Stock</font>") +
  theme(plot.title = element_markdown(), 
        text = element_text(size = 8)) +
  labs(x = "Stock Grant Value", y = "Total Yearly Compensation")

ggplot(data = salary) + 
  aes(x = stockgrantvalue, y = totalyearlycompensation) + 
  geom_point(aes(color = Race), size = .6) + 
  geom_smooth(aes(color = Race), method = "lm", se = FALSE, formula = 'y ~ x') + 
  ggtitle("<span style='font-size: 10pt;'>Salary vs Stock</font>") +
  theme(plot.title = element_markdown(), 
        text = element_text(size = 8)) +
  labs(x = "Stock Grant Value", y = "Total Yearly Compensation")
```

```{r echo=FALSE, fig.height=3, fig.width=8, fig.cap="Salary vs Bonus"}
# Bonus
ggplot(data = salary) + 
  aes(x = bonus, y = totalyearlycompensation) + 
  geom_point(aes(color = title_1), size = .6, alpha = 0.5) + 
  geom_smooth(aes(color = title_1), method = "lm", se = FALSE, formula = 'y ~ x') + 
  ggtitle("<span style='font-size: 10pt;'>Salary vs Bonus</font>") +
  theme(plot.title = element_markdown(), 
        text = element_text(size = 8)) +
  labs(x = "Bonus", y = "Total Yearly Compensation")

ggplot(data = salary) + 
  aes(x = bonus, y = totalyearlycompensation) + 
  geom_point(aes(color = Race), size = .6, alpha = 0.5) + 
  geom_smooth(aes(color = Race), method = "lm", se = FALSE, formula = 'y ~ x') + 
  ggtitle("<span style='font-size: 10pt;'>Salary vs Bonus</font>") +
  theme(plot.title = element_markdown(), 
        text = element_text(size = 8)) +
  labs(x = "Bonus", y = "Total Yearly Compensation")
```





