---
title: "MA678 Final Project"
author: "Yin Xu"
date: "2022-11-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, warning=FALSE, message=FALSE, include=FALSE}
library(ggplot2)
library(ggthemes)
library(Matrix)
library(lme4)
library(pbkrtest)
library(tidyr)
library(dplyr)
library(corrplot)
library(PerformanceAnalytics)
library(lmerTest)
library(lattice)
```


## Abstract

In most universities, AI, Statistics and Data Science are some of the most popular majors, plenty of students choose these majors since if they have the Data Science background, they may find jobs easily after university. According to this issue, there is a dataset from data science jobs shows that the titles, locations and other variables which effect the salaries. To discuss this topic, I build a multilevel model with group `title` and `race` The report includes 4 main parts: introduction, method, result and discussion.

## Introduction

As we always know, the job titles, locations of companies, experience level and other factors usually influence the amount of a person's salary. However, after COVID-19 came into everyone's life, some of jobs became remote to avoid direct contact, therefore, remote jobs may become one factor which influence the salaries. Remote work may decrease the salaries, because the efficiency will be reduced; sometimes, because of the extension of the work hour, the salaries will be increased. Besides, some small companies prefer to attract talents with high wages; at the same time, large companies have complete talents and do not care about the whereabouts of employees with low salaries.

Therefore, multilevel model is needed to discover the effect of the factors. I divide the factors to two group: fixed effect factors(e.g. education, bonus, location) and random effect factors(title, race).

\newpage

## Method

### Data Cleaning

The data is from kaggle open dataset(https://www.kaggle.com/datasets/jackogozaly/data-science-and-stem-salaries).

This data includes the variables of salaries from 2020 to 2021 around the world. 

First of all, there are "NA"s in the data, cleaning data is essential. Then I want to focus on the jobs in the United States, so the observations outside of USA should be dropped. Besides, I dropped several observations with super high wages (`total yearly compensation`) which is not fit for the plots. 

Secondly, there are 15 unique titles in `title`, but I only want to keep the position of the titles (e.g. manager, engineer, sales and so on). Therefore, I extracted the last word from `title` and added a new column named `title_1`, this column will be used as title for further analysis.

Additionally, in this data, `gender` and `Education` are two variables in characters, to make my exploratory data analysis more reasonable and easy to read, I changed `gender` to numeric values 0, 1, 2 which means other, female and male; for `Education`, 0 means "High School", 1 means "Some College", 2 means "Bachelor's Degree", 3 means "Master's Degree", and 4 means "PhD".

```{r}
salaries <- read.csv("Levels_Fyi_Salary_Data.csv")
salary <- salaries %>% drop_na()
```

```{r, warning=FALSE}
#unique(salary$title)
salary$title_1 <- gsub(".+\\s(.+)$", "\\1", salary$title)
salary$state <- gsub(".+\\s(.+)$", "\\1", salary$location)
#drop

# only USA & <= 1,000,000
salary <- salary %>% 
  filter(!grepl("[a-z]", state)) %>%
  filter(totalyearlycompensation <= 1000000) %>%
  separate(timestamp, c("timestamp", "time"), sep = " (?=[^ ]+$)")

salary$timestamp <- as.Date(salary$timestamp, "%m/%d/%Y")

salary <- salary[-c(4, 10, 15, 17:28)]

salary$Education[salary$Education == "Highschool"] <- 0
salary$Education[salary$Education == "Some College"] <- 1
salary$Education[salary$Education == "Bachelor's Degree"] <- 2
salary$Education[salary$Education == "Master's Degree"] <- 3
salary$Education[salary$Education == "PhD"] <- 4
salary$Education <- as.numeric(salary$Education)

salary$gender[salary$gender == "Other"] <- 0
salary$gender[salary$gender == "Female"] <- 1
salary$gender[salary$gender == "Male"] <- 2
salary$gender <- as.numeric(salary$gender)
```

After data cleaning, the following is the chart of variables which are used in this report:
`timestamp`               : When the data was recorded.
`company`                 : Company names.
`title`                   : Job title.
`location`                : Job location.
`cityid`                  : City ID of the location.
`totalyearlycompensation` : Total yearly compensation which is added by `basesalary`, `stockgrantvalue`and `bonus`.
`basesalary`              : Base salary of the year.
`stockgrantvalue`         : Stock grant value.
`bonus`                   : Bonus of the year.
`yearsofexperience`       : Year of experience of Data Science jobs.
`yearsatcompany`          : Year of experience at said company.
`gender`                  : Gender of the observations.
`race`                    : Race of the observations.
`Education`               : Education background.

\newpage

###Exploratory Data Analysis

By the data after cleaning, I got 16937 observations and 14 variables, and I set `total yearly compensation` as the output of the analysis; the other 13 variables will be discussed in following analysis to know if they influence the salaries of these Data Science jobs. To figure out the relationships between the variables and build a model, I select `title` and `race` as two groups.

# Year of experience
# experience yrs vs job title (y)
# experience yrs vs race

```{r}
ggplot(data = salary) + 
  aes(x = yearsofexperience, y = totalyearlycompensation) + 
  geom_jitter(aes(color = title_1), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = title_1), method = "lm", se = FALSE, formula = 'y ~ x') + 
  labs(title = "Salary vs Year of Experience", x = "Experience Years", y = "Total Yearly Compensation") 

ggplot(data = salary) + 
  aes(x = yearsofexperience, y = totalyearlycompensation) + 
  geom_jitter(aes(color = Race), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = Race), method = "lm", se = FALSE, formula = 'y ~ x') + 
  labs(title = "Salary vs Year of Experience", x = "Experience Years", y = "Total Yearly Compensation")
```

# Years at the Company
# yrs at company vs job title (y)
# yrs at company vs race

```{r}
ggplot(data = salary) + 
  aes(x = yearsatcompany, y = totalyearlycompensation) + 
  geom_jitter(aes(color = title_1), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = title_1), method = "lm", se = FALSE, formula = 'y ~ x') + 
  labs(title = "Salary vs Year at the Company", x = "Years at the Company", y = "Total Yearly Compensation") 

ggplot(data = salary) + 
  aes(x = yearsatcompany, y = totalyearlycompensation) + 
  geom_jitter(aes(color = Race), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = Race), method = "lm", se = FALSE, formula = 'y ~ x') + 
  labs(title = "Salary vs Year at the Company", x = "Years at the Company", y = "Total Yearly Compensation")
```

# Education
# education vs job title (y)
# education vs race

manager & marketing with PhD degree

```{r}
ggplot(data = salary) + 
  aes(x = Education, y = totalyearlycompensation) + 
  geom_jitter(aes(color = title_1), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = title_1), method = "lm", se = FALSE, formula = 'y ~ x') + 
  labs(title = "Salary vs Education", x = "Education", y = "Total Yearly Compensation")

ggplot(data = salary) + 
  aes(x = Education, y = totalyearlycompensation) + 
  geom_jitter(aes(color = Race), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = Race), method = "lm", se = FALSE, formula = 'y ~ x') + 
  labs(title = "Salary vs Education", x = "Education", y = "Total Yearly Compensation")
```

# Location
# location vs job title (y)
# location vs race

```{r}
ggplot(data = salary) + 
  aes(x = cityid, y = totalyearlycompensation) + 
  geom_jitter(aes(color = title_1), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = title_1), method = "lm", se = FALSE, formula = 'y ~ x') + 
  labs(title = "Salary vs Location", x = "City in USA", y = "Total Yearly Compensation")

ggplot(data = salary) + 
  aes(x = cityid, y = totalyearlycompensation) + 
  geom_jitter(aes(color = Race), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = Race), method = "lm", se = FALSE, formula = 'y ~ x') + 
  labs(title = "Salary vs Location", x = "City in USA", y = "Total Yearly Compensation")
```


# Gender
# gender vs job title
# gender vs race (y)

```{r, warning=FALSE}
ggplot(data = salary) + 
  aes(x = gender, y = totalyearlycompensation) + 
  geom_jitter(aes(color = title_1), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = title_1), method = "lm", se = FALSE, formula = 'y ~ x') + 
  labs(title = "Salary vs Gender", x = "Gender", y = "Total Yearly Compensation")

ggplot(data = salary) + 
  aes(x = gender, y = totalyearlycompensation) + 
  geom_jitter(aes(color = Race), size = .6, alpha = 0.1) + 
  geom_smooth(aes(color = Race), method = "lm", se = FALSE, formula = 'y ~ x') + 
  labs(title = "Salary vs Gender", x = "Gender", y = "Total Yearly Compensation")
```




## Correlation

```{r, echo=FALSE, include=FALSE}
cor <- salary[, c(7:11, 13)]
cor$yearsofexperience <- log(salary$yearsofexperience+1)
cor$yearsatcompany <- log(salary$yearsatcompany+1)
cor$basesalary <- log(salary$basesalary+1)
cor$stockgrantvalue <- log(salary$stockgrantvalue+1)
cor$bonus <- log(salary$bonus+1)
cor$gender <- log(salary$gender + 1)
cor$cityid <- log(salary$cityid+1)
cor$education <- salary$Education
cor$title <- salary$title_1
cor$race <- salary$Race
cor$salary <- log(salary$totalyearlycompensation + 1)
```

```{r, echo=FALSE, warning=FALSE, fig.cap="Correlation Matrix"}
chart.Correlation(cor[,1:7], histogram=TRUE, pch="+")
```


## Multilevel Model

```{r, eval=FALSE, warning=FALSE}
model <- lmer(salary ~ yearsofexperience + yearsatcompany + cityid + education + gender
              + (1 + yearsofexperience + yearsatcompany + cityid + education | title) 
              + (1 | race), 
              data = cor, REML = F)
summary(model)
anova(model)
```

## Model checking
```{r}
qqmath(model)
ranef(model)
```

\newpage


## Result



\newpage

## Appendix

# Time
# time vs job title
# time vs race

```{r include=FALSE}
ggplot(data = salary) + 
  aes(x = timestamp, y = totalyearlycompensation) + 
  geom_point(aes(color = title_1), size = .6) + 
  geom_smooth(aes(color = title_1), method = "lm", se = FALSE, formula = 'y ~ x') + 
  labs(title = "Salary vs Time", x = "Time stamp", y = "Total Yearly Compensation")

ggplot(data = salary) + 
  aes(x = timestamp, y = totalyearlycompensation) + 
  geom_point(aes(color = Race), size = .6) + 
  geom_smooth(aes(color = Race), method = "lm", se = FALSE, formula = 'y ~ x') + 
  labs(title = "Salary vs Time", x = "Time stamp", y = "Total Yearly Compensation")
```


# Stock Grant Value
# stock grant value vs job title 
# stock grant value vs race

```{r include=FALSE}
ggplot(data = salary) + 
  aes(x = stockgrantvalue, y = totalyearlycompensation) + 
  geom_point(aes(color = title_1), size = .6) + 
  geom_smooth(aes(color = title_1), method = "lm", se = FALSE, formula = 'y ~ x') + 
  labs(title = "Salary vs Stock", x = "Stock Grant Value", y = "Total Yearly Compensation")

ggplot(data = salary) + 
  aes(x = stockgrantvalue, y = totalyearlycompensation) + 
  geom_point(aes(color = Race), size = .6) + 
  geom_smooth(aes(color = Race), method = "lm", se = FALSE, formula = 'y ~ x') + 
  labs(title = "Salary vs Stock", x = "Stock Grant Value", y = "Total Yearly Compensation")
```

# Bonus
# bonus vs job title
# stock grant value vs race

```{r}
ggplot(data = salary) + 
  aes(x = bonus, y = totalyearlycompensation) + 
  geom_point(aes(color = title_1), size = .6, alpha = 0.5) + 
  geom_smooth(aes(color = title_1), method = "lm", se = FALSE, formula = 'y ~ x') + 
  labs(title = "Salary vs Bonus", x = "Bonus", y = "Total Yearly Compensation")

ggplot(data = salary) + 
  aes(x = bonus, y = totalyearlycompensation) + 
  geom_point(aes(color = Race), size = .6, alpha = 0.5) + 
  geom_smooth(aes(color = Race), method = "lm", se = FALSE, formula = 'y ~ x') + 
  labs(title = "Salary vs Bonus", x = "Bonus", y = "Total Yearly Compensation")
```





